Storm surge modelling in the Strait of Georgia

### 2014-12-04 RoughWork/manuscript.tex
Consistent labels
new table with averages

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statisitics, along with the mean observed values and mean modelled valuesm are reported in Table \ref{tab:statistics} for each hindcast. Further, comparisons between the maximum observed and modelled water level and residual are typically reported in the text.

The five hindcast simulations have been used to assess the model's skill and to learn more about how the storm surge propagates through the Salish Sea domain.  First, a general assessment of the model's skill is presented in two hindcasts during Feb 2006 and Nov 2006 in section \ref{sec:hind}. The Feb 2006 case is also used to determine the importance of modelling tidal forcing with the surge propogation. Second, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed for the Feb 2006 case in section \ref{sec:factors}. Third, the spatial variability of the surge as it propagates through the domain is examined for a Dec 2006 storm surge in section \ref{sec:spatial}. Fourth, a high wind event that did not result in a large surge or flooding for a Nov 2009 case is discussed in section \ref{sec:wind}. Finally, comparisons between hindcasts using high resolution and low resolution atmospheric forcing is provided in section \ref{sec:res}.

@@

Point Atkinson & 14.0           &  3.44                & 3.55                 &   0.012    & 0.995  &  6.1           &  0.313                 & 0.331                  &  0.237     & 0.914 \\
Victoria       &  9.9           &  2.26                & 2.24                 &   0.023    & 0.994  &  4.8           &  0.323                 & 0.293                  &  0.245     & 0.899 \\
Patricia Bay   & 11.2           &  2.60                & 2.65                 &   0.020    & 0.993  &  6.0           &  0.277                 & 0.290                  &  0.272     & 0.901 \\
Campbell River & 13.5           &  3.22                & 3.27                 &   0.019    & 0.995  &  4.9           &  0.326                 & 0.314                  &  0.217     & 0.924 \\

@@

@@

\caption{Evaluation of model performance for all hindcasts presented. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and residuals ($\delta$). Statisitics are calculated over the time period displayed in esch simulation's corresponding figure.}

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_factors.pdf}
\caption{Comparison of the modelled residual for four simulations in Feb 2006: A simulation with all forcing (solid black), a simulation without atmospheric forcing (solid grey), a simulation without the sea surface height anomaly (dashed black), and a simulation with only wind forcing, i.e.\ no sea surface height anomaly or pressure (dashed grey). }
\label{fig:factors}
\end{figure}

@@

Both of the simulations agree well with the observations (\ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly underpredicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Hindcasts of five significant events in February 2006, November 2006, December, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

Comparisons between high resolution and low resolution atmospheric forcing further suggest that the local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately in reproducing surge amplitudes as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy for storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of out testing was performed with the low resolution model. Additionally, the sea surface height anomaly from Neah Bay, forecasted by the National Oceanic and Atmospheric Administration (NOAA), will be used in future real-time water level predictions during the storm surge season. 
Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.

>>>>

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin-up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents. Forcing the model with only eight tidal constituents leads to an error in modelled sea surface height due to the omission of the next leading order tidal constituents such as J$_1$. Comparisons between full tidal predictions and tidal predictions using eight constituents suggest that this error can approach 40 cm during times of high tidal range, and in particular, during the storm surge season Nov-Feb. As such, storm surge model predictions were corrected by accounting for this error. Additionally, since the sea surface height is calculated about a zero reference level, the modelled water level is determined by adding a long term mean sea level at the specified location to the modelled sea surface height. All times are reported in UTC.

@@

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statistics, along with the mean observed values and mean modelled values are reported in Table \ref{tab:statistics} for each hindcast. Further, comparisons between the maximum observed and modelled water level and residual are typically reported in the text.

The five hindcast simulations have been used to assess the model's skill and to learn more about how the storm surge propagates through the Salish Sea domain.  First, a general assessment of the model's skill is presented in two hindcasts during Feb 2006 and Nov 2006 in section \ref{sec:hind}. The Feb 2006 case is also used to determine the importance of modelling tidal forcing with the surge propagation. Second, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed for the Feb 2006 case in section \ref{sec:factors}. Third, the spatial variability of the surge as it propagates through the domain is examined for a Dec 2006 storm surge in section \ref{sec:spatial}. Fourth, a high wind event that did not result in a large surge or flooding for a Nov 2009 case is discussed in section \ref{sec:wind}. Finally, comparisons between hindcasts using high resolution (HRDPS) and low resolution (CGRF) atmospheric forcing is provided in section \ref{sec:res}.

@@

Point Atkinson & 14.0           &  3.43                & 3.54                 &   0.012    & 0.995  &  5.8           &  0.295                 & 0.316                  &  0.200     & 0.926 \\
Victoria       &  9.9           &  2.25                & 2.23                 &   0.024    & 0.988  &  4.5           &  0.312                 & 0.280                  &  0.211     & 0.911 \\
Patricia Bay   & 10.3           &  2.59                & 2.64                 &   0.019    & 0.997  &  5.6           &  0.265                 & 0.275                  &  0.230     & 0.918 \\
Campbell River & 13.6           &  3.22                & 3.26                 &   0.019    & 0.991  &  4.5           &  0.313                 & 0.297                  &  0.182     & 0.934 \\

@@

Averages       &                &                      &                      &            &        &                &                        &                        &           &         \\                
Point Atkinson &	10.8	& 3.42	& 3.48	& 0.018	& 0.995	& 7.2	& 0.274	& 0.270	& 0.166	 & 0.939 \\
Victoria       &	10.2	& 2.19	& 2.15	& 0.041	& 0.986	& 5.9	& 0.254	& 0.238	& 0.168	 & 0.935 \\
Patricia Bay   &	10.6	& 2.59	& 2.58	& 0.030	& 0.991	& 6.6	& 0.243	& 0.236	& 0.169	 & 0.934 \\
Campbell River &	11.9	& 3.18	& 3.22	& 0.028	& 0.992	& 6.3	& 0.260	& 0.255	& 0.166  & 0.938 \\

@@

\caption{Evaluation of model performance for all hindcasts presented. Mean absolute error ($\bar{e}$), observed mean (subscript o), modelled mean (subscript m), $\gamma^2$, and the Willmott skill score ($WS$) are provided for water level predictions ($\eta$) and residuals ($\delta$). Statistics are calculated over the time period displayed in each simulation's corresponding figure (three or four days). For the Nov 2006 case, the statistics were calculated over a three day period. The average values were calculated over all of the hindcasts, excluding the Dec 2012 HRDPS simulation. }

@@

The contribution to the surge due to atmospheric forcing appears to be geographically dependent, its removal leading to a drop of about 2.5 cm in the maximum surge at Point Atkinson and 2.3 cm at Campbell River. Interestingly, the surge amplitude increases by 4.8 cm at Victoria and 3.6 cm at Patricia Bay when atmospheric forces are neglected. The southeasterly winds over the Strait of Georgia could be acting to push water away from the southern tip of Vancouver Island, resulting in a slight decrease in the surge amplitude at Victoria and Patricia Bay when wind effects are included in the model. Additionally, the surge amplitude decreases by a few centimetres if the inverse barometer effect due to atmospheric pressure is not included, in contrast to other findings that reported the inverse barometer effect contributes up to $^1/_3$ of the total surge amplitude \citep{murty1995storm}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_factors.pdf}
\caption{Comparison of the modelled residual for four simulations in Feb 2006: A simulation with all forcing (solid black), a simulation without atmospheric forcing (dashed green), a simulation without the sea surface height anomaly (solid red), and a simulation with only wind forcing, i.e.\ no sea surface height anomaly or pressure (dashed blue). }
\label{fig:factors}
\end{figure}

@@

Both of the simulations agree well with the observations (\ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above 0.92 at these four locations. Further, the modelled residuals deviate only slightly from each other, most noticeable at Point Atkinson during the large wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher. Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was 68.3 cm. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparable to forcing with high resolution products. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain.

@@

Hindcasts of five significant events in February 2006, November 2006, December 2006, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases and generally lower skill scores for the residuals.  Further, in a single simulation, the skill scores are fairly consistent between the four stations examined. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

Comparisons between high resolution and low resolution atmospheric forcing further suggest that the local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately in reproducing surge amplitudes as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy for storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of our testing was performed with the low resolution model. Additionally, the sea surface height anomaly from Neah Bay, forecasted by the National Oceanic and Atmospheric Administration (NOAA), will be used in future real-time water level predictions during the storm surge season.

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Lastly, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas.
